---
title: "02_Data_Process"
knit: (function(input, encoding) {
  rmarkdown::render(input = input,
                    output_format = "html_document",
                    output_dir = here::here("Output", "HTML"),
                    knit_root_dir = rprojroot::find_rstudio_root_file())})
  # rmarkdown::render(input = input,
  #                   output_format = "pdf_document",
  #                   output_dir = here::here("Output", "PDF"),
  #                   knit_root_dir = rprojroot::find_rstudio_root_file())})
output:
  html_document: 
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document: 
    toc: yes
date: "2025-03-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r warning=FALSE, message=FALSE}
library(dplyr)
library(caret)
library(keras)
library(purrr)
library(ggplot2)
library(recipes)
library(reshape2)
library(ggridges)
library(gridExtra)
library(patchwork)
library(tidymodels)
library(FactoMineR)
```


## Load

Loading Pre-processed data provided

```{r}
## load
load(here::here("Data", "ProcessedData", "processed_AnalysisData.Rdata"))
```

```{r}
processed_data
```

```{r}
## variable names in data
names(processed_data)
```


## Data Preparation

1. We only focuses on the target strength data from the fish species, thus, we first only create a dataframe that contains the frequency response across all pings for fish.

```{r}
frequency_data <- (
  processed_data
  |> dplyr::select(c(1, 5, 59:484))
)
frequency_data |> head()
```


2. We next remove the fish `LT008` because it is missing all high frequency data from `F173` to `F258.5` for all pings, where results in a total of 31476 `NA` value and fill in values is not meaningful at all.

```{r}
frequency_data <- (
  frequency_data
  |> filter(fishNum != "LT008")
)
unique(frequency_data$fishNum)
```

3. We then remove the data columns `F90` to `F170` because they did not record in the original data, where all columns are `NA` values for all ping data and all fish.

```{r}
frequency_data <- frequency_data |> dplyr::select(-c(93:256))
```

4. Prepare class levels and number of classes in the data for model development use

```{r}
all_species <- unique(frequency_data$species)
num_classes <- length(all_species)
```

### Train and Test set by fishNum

We then want to create a train and test set by fish ID. We want to randomly choose 70% Train set and 30% Test set.

```{r}
set.seed(2453)

## unique fish id
unique_fishNum <- frequency_data |> distinct(fishNum, species)

## create stratified split
train_indices <- createDataPartition(
  unique_fishNum$species,
  p = 0.8,
  list = FALSE
)

## create train/test fish ID set
train_fish <- unique_fishNum[train_indices, ]
test_fish <- unique_fishNum[-train_indices, ]

## split into train/test set
train_data <- frequency_data |> filter(fishNum %in% train_fish$fishNum)
test_data <- frequency_data |> filter(fishNum %in% test_fish$fishNum)
dim(train_data)
dim(test_data)
```


### Prepare Data for different models

We want to create a classification system that using the following models:

1. Logistics regression (with PCA)
2. Random Forest (Raw data)
3. Deep Neural Network (Raw data)
4. LSTM (Raw Sequence data)
5. LSTM with attention layer (Raw Sequence data)
6. CNN (Raw Sequence data)

We need to prepare three type od data.
```{r}
## standardized data
preProc <- preProcess(train_data, method = c("center", "scale"))
train_std_data <- predict(preProc, train_data)
test_std_data <- predict(preProc, test_data)

## raw data
train_raw_data <- train_data
test_raw_data <- test_data

## create sequence data
create_sequences_data <- function(data) {
  
  ## group by fish
  grouped_data <- (
    data
    |> group_by(fishNum)
    |> mutate(n_pings = n())
    |> ungroup()
    |> group_split(fishNum)
  )
  
  # Extract metadata
  labels <- map_chr(grouped_data, ~ unique(.x$species))
  # labels <- unique(grouped_data$species)
  fishNums <- map_chr(grouped_data, ~ unique(.x$fishNum))
  # fishNums <- unique(grouped_data$fishNum)
  
  ## convert to list of matrices
  sequences <- (
    grouped_data
    |> map(~ {
      freq_cols <- (
        .x
        |> dplyr::select(-c(fishNum, species))
        |> dplyr::select(where(is.numeric))
      )
      as.matrix(freq_cols)
    })
  )
  
  return(list(
    fishNum = fishNums,
    species = labels,
    sequences = sequences
  ))
}
train_seq_data <- create_sequences_data(train_std_data)
test_seq_data <- create_sequences_data(test_std_data)
```


## Save

```{r}
# save.image(file = here::here("Output", "Data", "02_Data_Process.RData"))
```













