---
title: "02_Models"
knit: (function(input, encoding) {
  rmarkdown::render(input = input,
                    output_format = "html_document",
                    output_dir = here::here("Output", "HTML"),
                    knit_root_dir = rprojroot::find_rstudio_root_file())})
  # rmarkdown::render(input = input,
  #                   output_format = "pdf_document",
  #                   output_dir = here::here("Output", "PDF"),
  #                   knit_root_dir = rprojroot::find_rstudio_root_file())})
output:
  html_document: 
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document: 
    toc: yes
date: "2025-03-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r warning=FALSE, message=FALSE}
library(dplyr)
library(caret)
library(keras)
library(purrr)
library(ggplot2)
library(recipes)
library(reshape2)
library(ggridges)
library(gridExtra)
library(patchwork)
library(tidymodels)
library(FactoMineR)
```


## Load

Loading Pre-processed data provided

```{r}
## load
load(here::here("Data", "ProcessedData", "processed_AnalysisData.Rdata"))
```

```{r}
processed_data
```

```{r}
## variable names in data
names(processed_data)
```


## Data Preparation

1. We only focuses on the target strength data from the fish species, thus, we first only create a dataframe that contains the frequency response across all pings for fish.

```{r}
frequency_data <- (
  processed_data
  |> dplyr::select(c(1, 5, 59:481))
)
frequency_data |> head()
```


2. We next remove the fish `LT008` because it is missing all high frequency data from `F173` to `F258.5` for all pings, where results in a total of 31476 `NA` value and fill in values is not meaningful at all.

```{r}
frequency_data <- (
  frequency_data
  |> filter(fishNum != "LT008")
)
unique(frequency_data$fishNum)
```

3. We then remove the data columns `F90` to `F170` because they did not record in the original data, where all columns are `NA` values for all ping data and all fish.

```{r}
frequency_data <- frequency_data |> dplyr::select(-c(93:253))
```

4. Prepare class levels and number of classes in the data for model development use

```{r}
all_species <- unique(frequency_data$species)
num_classes <- length(all_species)
```

### Train and Test set by fishNum

We then want to create a train and test set by fish ID. We want to randomly choose 70% Train set and 30% Test set.

```{r}
set.seed(2453)

## unique fish id
unique_fishNum <- frequency_data |> distinct(fishNum, species)

## create stratified split
train_indices <- createDataPartition(
  unique_fishNum$species,
  p = 0.8,
  list = FALSE
)

## create train/test fish ID set
train_fish <- unique_fishNum[train_indices, ]
test_fish <- unique_fishNum[-train_indices, ]

## split into train/test set
train_data <- frequency_data |> filter(fishNum %in% train_fish$fishNum)
test_data <- frequency_data |> filter(fishNum %in% test_fish$fishNum)
dim(train_data)
dim(test_data)
```


### Prepare Data for different models

We want to create a classification system that using the following models:

1. Logistics regression (with PCA)
2. Random Forest (Raw data)
3. Deep Neural Network (Raw data)
4. LSTM (Raw Sequence data)
5. LSTM with attention layer (Raw Sequence data)
6. CNN (Raw Sequence data)

We need to prepare three type od data.
```{r}
## standardized data
preProc <- preProcess(train_data, method = c("center", "scale"))
train_std_data <- predict(preProc, train_data)
test_std_data <- predict(preProc, test_data)

## raw data
train_raw_data <- train_data
test_raw_data <- test_data

## create sequence data
create_sequences_data <- function(data) {
  
  ## group by fish
  grouped_data <- (
    data
    |> group_by(fishNum)
    |> mutate(n_pings = n())
    |> ungroup()
    |> group_split(fishNum)
  )
  
  # Extract metadata
  labels <- map_chr(grouped_data, ~ unique(.x$species))
  # labels <- unique(grouped_data$species)
  fishNums <- map_chr(grouped_data, ~ unique(.x$fishNum))
  # fishNums <- unique(grouped_data$fishNum)
  
  ## convert to list of matrices
  sequences <- (
    grouped_data
    |> map(~ {
      freq_cols <- (
        .x
        |> dplyr::select(-c(fishNum, species))
        |> dplyr::select(where(is.numeric))
      )
      as.matrix(freq_cols)
    })
  )
  
  return(list(
    fishNum = fishNums,
    species = labels,
    sequences = sequences
  ))
}
train_seq_data <- create_sequences_data(train_std_data)
test_seq_data <- create_sequences_data(test_std_data)
```



## Models

### Logistics Regression

```{r}
###############################################################################
# PCA-Logistic Regression Pipeline with 90% Variance Retention
# ------------------------------------------------------------
# This script:
# 1. Applies PCA to standardized frequency data
# 2. Retains components explaining ≥90% cumulative variance
# 3. Trains logistic regression on reduced dimensions
# 4. Evaluates performance on test set
###############################################################################

## fish level standard data
train_std_fish_level_data <- (
  train_std_data
  |> group_by(fishNum, species)
  |> summarise(across(starts_with("F"), mean))
  |> ungroup()
)

test_std_fish_level_data <- (
  test_std_data
  |> group_by(fishNum, species)
  |> summarise(across(starts_with("F"), mean))
  |> ungroup()
)

## Apply PCA
## define recipe
pca_recipe <- (
  recipe(species ~ ., data = train_std_fish_level_data)
  ## keeping fishNum but exclude from analysis
  |> update_role(fishNum, new_role = "ID")
  ## want PCs at least have 90% variance explained
  |> step_pca(
    all_numeric_predictors(),
    threshold = 0.9,
    options = list(center = FALSE, scale. = FALSE)
  )
)

## define model
## using multinomial regression
multinom_model <- (
  multinom_reg()
  |> set_engine("nnet")
  |> set_mode("classification")
)

## assembly workflow
pca_multinom_wf <- (
  workflow()
  |> add_recipe(pca_recipe)
  |> add_model(multinom_model)
)

## train model
fit_pca_multinom <- (
  pca_multinom_wf
  |> fit(data = train_std_fish_level_data)
)

## Variance Explanation Analysis
## extract PCA details
pca_multinom_res <- (
  extract_preprocessor(fit_pca_multinom)
  |> prep()
)

## extract variance explained metrics
pca_multinom_model <- pca_multinom_res$steps[[1]]$res
pca_multinom_variance <- pca_multinom_model$sdev^2
pca_multinom_variance_cum_variance <- cumsum(pca_multinom_variance / sum(pca_multinom_variance))

pca_multinom_variance_tbl <- (
  tibble(
    component = paste0("PC", seq_along(pca_multinom_variance_cum_variance)),
    cumulative_variance = pca_multinom_variance_cum_variance
  )
  |> filter(
    cumulative_variance <= 0.9 | row_number() == which.max(cumulative_variance >= 0.9)
  )
)

# print
cat("\nPCA Variance Explanation:\n")
print(pca_multinom_variance_tbl)
cat("Number of components retained:", nrow(pca_multinom_variance_tbl), "\n")
```


```{r}
## Model Predictions
## generate test results
predict_pca_multinom <- (
  test_std_fish_level_data
  |> bind_cols(
    predict(fit_pca_multinom, new_data = test_std_fish_level_data), ## compute class predictions
    predict(fit_pca_multinom, new_data = test_std_fish_level_data, type = "prob") ## compute probabilities
  )
  |> mutate(species = factor(species))
)

## performance metrics
model_metrics <- metric_set(accuracy, sensitivity, specificity)
performance_pca_multinom <- (
  predict_pca_multinom
  |> model_metrics(
    truth = species,
    estimate = .pred_class
  )
)

# Compute multiclass roc_auc separately with all probability columns
pca_multinom_roc_auc_res <- (
  predict_pca_multinom
  |> roc_auc(
    truth = species,
    c(.pred_lakeTrout, .pred_lakeWhitefish, .pred_smallmouthBass),
    estimator = "hand_till"
  )
)


## Model Performance
## confusion matrix
cat("\nConfusion Matrix:\n")
print(conf_mat(predict_pca_multinom, truth = species, estimate = .pred_class))

cat("\nModel Performance Metrics:\n")
print(performance_pca_multinom)

## PCA laoding contributions
pca_multinom_loadings <- pca_multinom_model$rotation[, "PC1"]
pca_multinom_loadings_contributions_tbl <- (
  tibble(
    Frequency = names(pca_multinom_loadings),
    Loading = pca_multinom_loadings,
    Absolute_Loading = abs(pca_multinom_loadings)
  )
  |> arrange(desc(Absolute_Loading))
)

head(pca_multinom_loadings_contributions_tbl, 20)
```


### Random Forest

```{r}
set.seed(2453)

## Random Forest
## define recipe
rf_recipe <- (
  recipe(species ~ ., data = train_raw_data)
  |> update_role(fishNum, new_role = "ID")
)

## define random forest model
rf_model <- (
  rand_forest(
    trees = tune(),
    mtry = tune(),
    min_n = tune()
  )
  |> set_engine("ranger", importance = "impurity")
  |> set_mode("classification")
)

## workflow assemble
rf_wf <- (
  workflow()
  |> add_recipe(rf_recipe)
  |> add_model(rf_model)
)

## tune hyperparameters
rf_folds <- vfold_cv(train_raw_data, v = 5)
rf_grid <- grid_regular(
  trees(range = c(100, 500)), # 300  600  900 1200 1500
  mtry(range = c(10, 50)),    # 10  32  55  77 100
  min_n(range = c(1, 20)),     # 1  5 10 15 20
  levels = 5
)

## took 6 hours to tune
# rf_tuned_model <- (
#   rf_wf
#   |> tune_grid(
#     resamples = rf_folds,
#     grid = rf_grid,
#     metrics = metric_set(accuracy)
#     )
#   )
# ## save to RDS
# save(rf_tuned_model, file = here::here("Output", "Data", "02_RF_Tuned_Models.rds"))
```


```{r}
## load tuned models
load(here::here("Output", "Data", "02_RF_Tuned_Models.rds"))
# Select the best hyperparameters
rf_best_parameters <- rf_tuned_model |> tune::select_best(metric = "accuracy")
rf_best_parameters

## finalize the best model for prediction
fit_rf_best <- (
  rf_wf
  |> finalize_workflow(rf_best_parameters)
  |> fit(train_raw_data)
)

## predict on the test data
predict_rf_best <- (
  test_raw_data
  |> bind_cols(
    predict(fit_rf_best, new_data = test_raw_data),
    predict(fit_rf_best, new_data = test_raw_data, type = "prob")
    )
  |> mutate(species = factor(species))
)

## performance metrics
rf_metric <- metric_set(accuracy)
performance_rf_best <- (
  predict_rf_best
  |> rf_metric(
    truth = species,
    estimate = .pred_class
  )
)


## Model Performance

## create tibble to store predictions for model performance comparison
rf_predictions <- tibble(
  fishNum = test_std_data$fishNum,
  species = test_std_data$species,
  pred_class = predict_rf_best$.pred_class
)

## count majority predictions on ping data for a single fish
## to determine the dnn prediction for this fish
rf_fish_level_pred <- (
  rf_predictions
  |> group_by(fishNum)
  |> summarise(
    pred_class = names(which.max(table(pred_class))),
    species = first(species)
  )
  |> mutate(
    pred_class = factor(pred_class, levels = all_species),
    species = factor(species, levels = all_species)
  )
)
rf_fish_level_pred

## compute and show ping predictions counts
rf_predictions_count <- (
  rf_predictions
  |> group_by(fishNum, species, pred_class)
  |> summarise(n = n(), .groups = "drop")
  |> pivot_wider(
    names_from = pred_class,
    values_from = n,
    values_fill = 0,
    names_prefix = "Pred_"
  )
)
rf_predictions_count$Pred_class <- rf_fish_level_pred$pred_class 
rf_predictions_count

# Compute accuracy
rf_accuracy <- (
  rf_fish_level_pred
  |> accuracy(truth = species, estimate = pred_class)
)

## confusion matrix
rf_conf_mat <- (
  rf_fish_level_pred
  |> conf_mat(truth = species, estimate = pred_class)
)

## Output
## confusion matrix
cat("\nConfusion Matrix:\n")
print(rf_conf_mat)

cat("\nModel Performance Metrics:\n")
print(rf_accuracy)
```



### Deep Nerual Network

```{r}
set.seed(2453)

#### some useful note
#### install.package("keras")
#### install.package("tensorflow)"
#### library(tensorflow)
#### install_tensorflow(version = "2.13.1")
#### for model to run properly

## Nerual Networks
## prepare data
train_dnn_data <- (
  train_std_data
  |> mutate(species = factor(species))
  |> select(-fishNum)
)

test_dnn_data <- (
  test_std_data
  |> mutate(species = factor(species))
  |> select(-fishNum)
)

## separate features and lables

## train
train_dnn_x <- (
  train_dnn_data
  |> dplyr::select(starts_with(("F")))
  |> as.matrix()
)

train_dnn_y <- (
  train_dnn_data
  |> pull(species)
  |> as.numeric() - 1
)

## test
test_dnn_x <- (
  test_dnn_data
  |> dplyr::select(starts_with(("F")))
  |> as.matrix()
)

test_dnn_y <- (
  test_dnn_data
  |> pull(species)
  |> as.numeric() - 1
)

## one-hot encode labels 
train_dnn_y <- to_categorical(train_dnn_y, num_classes = num_classes)
test_dnn_y <- to_categorical(test_dnn_y, num_classes = num_classes)

## DNN model
## define model with 3 hidden layers
dnn_model <- (
  keras_model_sequential()
  |> layer_dense(units = 128, activation = 'relu', input_shape = ncol(train_dnn_x))
  |> layer_dropout(rate = 0.2)
  |> layer_dense(units = 64, activation = 'relu')
  |> layer_dropout(rate = 0.2)
  |> layer_dense(units = 32, activation = "relu")
  |> layer_dense(units = num_classes, activation = "softmax")
)

## compile the model
(
  dnn_model
  |> compile(
    loss = "categorical_crossentropy",
    optimizer = optimizer_adam(learning_rate = 0.001),
    metrics = "accuracy"
  )
)

## create validation sets for model development
dnn_grouped_data <- (
  train_std_data
  |> mutate(group = fishNum)
  # |> group_by(group)
  # |> nest()
)
## create folds
dnn_folds <- group_initial_split(dnn_grouped_data, group = "fishNum", prop = 0.7)
train_dnn_folds <- (
  training(dnn_folds)
)
val_dnn_folds <- (
  testing(dnn_folds)
)

val_dnn_x <- (
  val_dnn_folds
  |> select(-fishNum)
  |> select(starts_with("F"))
  |> as.matrix()
)
val_dnn_y <- (
  val_dnn_folds
  |> mutate(species = factor(species))
  |> pull(species)
  |> as.numeric() - 1
)
val_dnn_y <- to_categorical(val_dnn_y, num_classes = num_classes)

## train model
dnn_history <- (
  dnn_model
  |> fit(
    x = train_dnn_x,
    y = train_dnn_y,
    epochs = 100,
    batch_size = 32,
    validation_data = list(val_dnn_x, val_dnn_y),
    verbose = 1
  )
)

## evaluate model
(
  dnn_model
  |> evaluate(test_dnn_x, test_dnn_y)
)

## predict
predict_dnn <- (
  dnn_model
  |> predict(test_dnn_x)
  |> k_argmax()
)
predict_dnn
```

```{r}
## Convert history to a tibble
dnn_history_df <- as.data.frame(dnn_history$metrics)
dnn_history_df$epoch <- seq(from = 1, to = 100)
dnn_history_df

## Plot loss
ggplot(dnn_history_df, aes(x = epoch)) +
  geom_line(aes(y = loss, color = "Training Loss")) +
  geom_line(aes(y = val_loss, color = "Validation Loss")) +
  labs(title = "Training vs. Validation Loss", y = "Loss", x = "Epoch") +
  theme_minimal()

## Plot accuracy
ggplot(dnn_history_df, aes(x = epoch)) +
  geom_line(aes(y = accuracy, color = "Training Accuracy")) +
  geom_line(aes(y = val_accuracy, color = "Validation Accuracy")) +
  labs(title = "Training vs. Validation Accuracy", y = "Accuracy", x = "Epoch") +
  theme_minimal()
```

```{r}
## Model Performance
## convert prediction to numeric values
dnn_pred_classes <- as.numeric(predict_dnn$numpy())

## create tibble to store predictions for model performance comparison
dnn_predictions <- tibble(
  fishNum = test_std_data$fishNum,
  species = test_std_data$species,
  pred_class = dnn_pred_classes
)

## count majority predictions on ping data for a single fish
## to determine the dnn prediction for this fish
dnn_fish_level_pred <- (
  dnn_predictions
  |> group_by(fishNum)
  |> summarise(
    pred_class = as.integer(names(which.max(table(pred_class)))),
    species = first(species)
  )
  |> ungroup()
  |> mutate(
    pred_class = case_when(
      pred_class == 0 ~ "lakeTrout",
      pred_class == 1 ~ "lakeWhitefish",
      pred_class == 2 ~ "smallmouthBass",
      TRUE ~ NA_character_ ## set default case for any other values
    ),
    species = factor(species, levels = all_species),
    pred_class = factor(pred_class, levels = all_species)
  )
)

## compute and show ping predictions counts
dnn_predictions_count <- (
  dnn_predictions
  |> group_by(fishNum, species, pred_class)
  |> summarise(n = n(), .groups = "drop")
  |> pivot_wider(
    names_from = pred_class,
    values_from = n,
    values_fill = 0,
    names_prefix = "Pred_"
  )
) 
dnn_predictions_count$Pred_class <- dnn_fish_level_pred$pred_class 

dnn_predictions_count

# Compute accuracy
dnn_accuracy <- (
  dnn_fish_level_pred
  |> accuracy(truth = species, estimate = pred_class)
)

## confusion matrix
dnn_conf_mat <- (
  dnn_fish_level_pred
  |> conf_mat(truth = species, estimate = pred_class)
)
  
## Output
## confusion matrix
cat("\nConfusion Matrix:\n")
print(dnn_conf_mat)

cat("\nModel Performance Metrics:\n")
print(dnn_accuracy)
```


### LSTM

```{r}
## LSTM models

create_pad_sequence <- function(m, max_length) {
  # Validate input is a matrix
  if (!is.matrix(m)) {
    stop("Input must be a matrix")
  }
  
  n_rows <- nrow(m)
  
  if (n_rows < max_length) {
    # Pad with zeros
    pad_rows <- max_length - n_rows
    padded <- matrix(0, nrow = pad_rows, ncol = ncol(m))
    rbind(m, padded)
  } else if (n_rows > max_length) {
    # Truncate to max_length
    m[1:max_length, , drop = FALSE]
  } else {
    # Return as-is
    m
  }
}

## prepare data
## want to convert list strcture to 3D array [fish, pings, features]

## find max pings per fish
train_ping_max_length <- max(map_dbl(train_seq_data$sequences, ~ dim(.)[1]))
test_ping_max_length <- max(map_dbl(test_seq_data$sequences, ~ dim(.)[1]))
lstm_ping_max_length <- max(train_ping_max_length, test_ping_max_length)
# train_ping_max_length <- 800
# test_ping_max_length <- 800
## number of frequency bands
train_num_features <- dim(train_seq_data$sequences[[1]])[2]
test_num_features <- dim(test_seq_data$sequences[[1]])[2]

## create 3d array
train_seq_array <- (
  train_seq_data$sequences
  |> map(function(x) create_pad_sequence(x, lstm_ping_max_length))
  |> unname()
  |> abind::abind(along = 3)
  |> aperm(c(3, 1, 2))
)

test_seq_array <- (
  test_seq_data$sequences
  |> map(function(x) create_pad_sequence(x, lstm_ping_max_length))
  |> unname()
  |> abind::abind(along = 3)
  |> aperm(c(3, 1, 2))
)

## define train sets
train_lstm_y <- (
  train_seq_data$species
  |> factor(levels = all_species)
  |> as.numeric() - 1
)

train_lstm_y <- to_categorical(train_lstm_y, num_classes = num_classes)

## define test sets
test_lstm_y <- (
  test_seq_data$species
  |> factor(levels = all_species)
  |> as.numeric() - 1
)

test_lstm_y <- to_categorical(test_lstm_y, num_classes = num_classes)
```


```{r}
## define model with 3 hidden layers
lstm_model <- (
  keras_model_sequential()
  ## mask layer to ignore padded values with 0
  |> layer_masking(
    mask_value = 0,
    input_shape = c(lstm_ping_max_length, train_num_features), ## (time_steps, features)
  )
  # |> layer_batch_normalization()
  # |> layer_lstm(
  #   units = 256,
  #   return_sequences = TRUE
  # )
  # |> layer_dropout(rate = 0.2)
  
  # |> layer_lstm(
  #   units = 64,
  #   kernel_regularizer = regularizer_l2(0.0001),
  #   return_sequences = TRUE
  # )
  # |> layer_dropout(rate = 0.2)
  # |> layer_lstm(
  #   units = 64,
  #   # kernel_regularizer = regularizer_l2(0.0001),
  #   return_sequences = TRUE
  # )
  # |> layer_dropout(rate = 0.3)
  |> layer_lstm(units = 64)
                # kernel_regularizer = regularizer_l2(0.0001))
  |> layer_dropout(rate = 0.4)
  ## classification layers
  |> layer_dense(units = 32, activation = "relu")
  |> layer_dropout(rate = 0.4)
  # |> layer_dense(units = 16, activation = "relu")
  |> layer_dense(units = num_classes, activation = "softmax")
)

lstm_model <- (
  keras_model_sequential()
  |> layer_masking(
    mask_value = 0,
    input_shape = c(lstm_ping_max_length, train_num_features)
  )
  |> layer_lstm(
    units = 128,  # Slightly larger first layer to capture the 263 features
    return_sequences = TRUE,  # Return sequences for stacked LSTM
    kernel_regularizer = regularizer_l2(0.001)
  )
  |> layer_dropout(rate = 0.4)
  |> layer_lstm(
    units = 64,  # Second LSTM layer to capture higher-level patterns
    kernel_regularizer = regularizer_l2(0.001)
  )
  |> layer_dropout(rate = 0.4)
  |> layer_dense(
    units = 32, 
    activation = "relu",
    kernel_regularizer = regularizer_l2(0.001)
  )
  |> layer_dropout(rate = 0.4)
  |> layer_dense(units = 3, activation = "softmax")
)

## compile model
(
  lstm_model
  |> compile(
    loss = "categorical_crossentropy",
    optimizer = optimizer_adam(learning_rate = 0.005),
    # optimizer = optimizer_adam(learning_rate = 0.001, weight_decay = 0.0005),
    metrics = "accuracy"
  )
)

## create validation sets for model development
train_fishNum <- unique(train_seq_data$fishNum)
# lstm_val_fishNum <- sample(train_fishNum,
#                            size = floor(0.2 * length(train_fishNum)))
lstm_val_fishNum <- (
  train_std_data
  |> distinct(species, fishNum)  # Get unique species-fishNum combinations
  |> group_by(species)
  |> slice_sample(prop = 0.2)
  |> pull(fishNum)
)

## split data using fishNum
train_lstm_idx <- which(!train_seq_data$fishNum %in% lstm_val_fishNum)
val_lstm_idx <- which(train_seq_data$fishNum %in% lstm_val_fishNum)


class_counts <- table(train_seq_data$species)
total <- sum(class_counts)
class_weights <- list(
  "0" = total / (num_classes * class_counts[1]),  # LakeTrout
  "1" = total / (num_classes * class_counts[2]),  # LakeWhitefish
  "2" = total / (num_classes * class_counts[3])   # smallmouthBass
)

## sample weights
# Create weights based on sequence length (giving more weight to shorter sequences)
sequence_lengths <- map_dbl(train_seq_data$sequences, ~ nrow(.))

# Option 1: Inverse weighting (shorter sequences get more weight)
sample_weights <- 1 / (sequence_lengths + 1)  # Adding 1 to avoid division by zero

# Option 2: Normalize weights to sum to n_samples
sample_weights <- sample_weights * (length(sample_weights) / sum(sample_weights))

# Select only weights for training samples
train_sample_weights <- sample_weights[train_lstm_idx]

## train model
lstm_history <- (
  lstm_model
  |> fit(
    x = train_seq_array[train_lstm_idx, , ],
    y = train_lstm_y[train_lstm_idx, ],
    epochs = 100,
    batch_size = 16,
    validation_data = list(
      train_seq_array[val_lstm_idx, , ],
      train_lstm_y[val_lstm_idx, ]
    ),
    callbacks = list(
      callback_early_stopping(patience = 15),
      callback_reduce_lr_on_plateau(factor = 0.5, patience = 5)
    ),
    # callbacks = callbacks_list,
    # sample_weight = train_sample_weights,
    class_weight = class_weights,
    verbose = 1
  )
)
```


```{r}
set.seed(2453)

## LSTM models

## prepare data
## want to convert list strcture to 3D array [fish, pings, features]

## find max pings per fish
train_ping_max_length <- max(map_dbl(train_seq_data$sequences, ~ dim(.)[1]))
test_ping_max_length <- max(map_dbl(test_seq_data$sequences, ~ dim(.)[1]))
lstm_ping_max_length <- max(train_ping_max_length, test_ping_max_length)
# train_ping_max_length <- 800
# test_ping_max_length <- 800
## number of frequency bands
train_num_features <- dim(train_seq_data$sequences[[1]])[2]
test_num_features <- dim(test_seq_data$sequences[[1]])[2]

## padding dunction
# create_pad_sequence <- function(m, max_length) {
#   # rbind(m,
#   #       matrix(0,
#   #              nrow = max_length - nrow(m),
#   #              ncol = ncol(m)))
#   
#   if (nrow(m) < max_length) {
#     padded <- matrix(0,
#                      nrow = max_length - nrow(m),
#                      ncol = ncol(m))
#     return(rbind(m, padded))
#   } else if (is.list(m)) {
#     # If m is a list, recursively apply the function to each element
#     return(lapply(m, function(x) create_pad_sequence(x, max_length)))
#   } else {
#     return(m[1:max_length, , drop = FALSE])
#   }
  
  # if (is.matrix(m) || is.data.frame(m)) {
  #   if (nrow(m) < max_length) {
  #     # If sequence is shorter than max_length, pad it
  #     padded <- matrix(0,
  #                      nrow = max_length - nrow(m),
  #                      ncol = ncol(m))
  #     return(rbind(m, padded))
  #   } else {
  #     # If sequence is longer or equal to max_length, truncate it
  #     return(m[1:max_length, , drop = FALSE])
  #   }
  # } else if (is.list(m)) {
  #   # If m is a list, recursively apply the function to each element
  #   return(lapply(m, function(x) create_pad_sequence(x, max_length)))
  # } else {
  #   stop("Input must be a matrix, data frame, or list of matrices")
  # }
# }

create_pad_sequence <- function(m, max_length) {
  # Validate input is a matrix
  if (!is.matrix(m)) {
    stop("Input must be a matrix")
  }
  
  n_rows <- nrow(m)
  
  if (n_rows < max_length) {
    # Pad with zeros
    pad_rows <- max_length - n_rows
    padded <- matrix(0, nrow = pad_rows, ncol = ncol(m))
    rbind(m, padded)
  } else if (n_rows > max_length) {
    # Truncate to max_length
    m[1:max_length, , drop = FALSE]
  } else {
    # Return as-is
    m
  }
}

## create 3d array
train_seq_array <- (
  train_seq_data$sequences
  |> map(function(x) create_pad_sequence(x, lstm_ping_max_length))
  |> unname()
  |> abind::abind(along = 3)
  |> aperm(c(3, 1, 2))
)

test_seq_array <- (
  test_seq_data$sequences
  |> map(function(x) create_pad_sequence(x, lstm_ping_max_length))
  |> unname()
  |> abind::abind(along = 3)
  |> aperm(c(3, 1, 2))
)

## define train sets
train_lstm_y <- (
  train_seq_data$species
  |> factor(levels = all_species)
  |> as.numeric() - 1
)

train_lstm_y <- to_categorical(train_lstm_y, num_classes = num_classes)

## define test sets
test_lstm_y <- (
  test_seq_data$species
  |> factor(levels = all_species)
  |> as.numeric() - 1
)

test_lstm_y <- to_categorical(test_lstm_y, num_classes = num_classes)

## define model with 3 hidden layers
lstm_model <- (
  keras_model_sequential()
  ## mask layer to ignore padded values with 0
  |> layer_masking(
    mask_value = 0,
    input_shape = c(lstm_ping_max_length, train_num_features), ## (time_steps, features)
  )
  |> layer_batch_normalization()
  # |> layer_lstm(
  #   units = 256,
  #   return_sequences = TRUE
  # )
  # |> layer_dropout(rate = 0.2)
  
  # |> layer_lstm(
  #   units = 64,
  #   kernel_regularizer = regularizer_l2(0.0001),
  #   return_sequences = TRUE
  # )
  # |> layer_dropout(rate = 0.2)
  |> layer_lstm(
    units = 128,
    kernel_regularizer = regularizer_l2(0.0001),
    return_sequences = TRUE
  )
  |> layer_dropout(rate = 0.3)
  |> layer_lstm(units = 64,
                kernel_regularizer = regularizer_l2(0.0001))
  |> layer_dropout(rate = 0.3)
  ## classification layers
  |> layer_dense(units = 64, activation = "relu")
  |> layer_dense(units = 3, activation = "softmax")
)

# lstm_model <- ( keras_model_sequential() |> layer_masking(mask_value = 0, input_shape = c(lstm_ping_max_length, train_num_features)) |> layer_batch_normalization() |> layer_bidirectional(layer_lstm(units = 64, return_sequences = TRUE, kernel_regularizer = regularizer_l2(0.001))) |> layer_dropout(rate = 0.3) |> layer_lstm(units = 32, kernel_regularizer = regularizer_l2(0.001)) |> layer_dropout(rate = 0.3) |> layer_dense(units = 32, activation = "relu", kernel_regularizer = regularizer_l1_l2(l1 = 0.0001, l2 = 0.001)) |> layer_dense(units = 3, activation = "softmax") )

## compile model
(
  lstm_model
  |> compile(
    loss = "categorical_crossentropy",
    optimizer = optimizer_adam(learning_rate = 0.005),
    # optimizer = optimizer_adam(learning_rate = 0.001, weight_decay = 0.0005),
    # optimizer = optimizer_rmsprop(learning_rate = 0.001, rho = 0.9),
    # optimizer_adadelta(learning_rate = 1.0, rho = 0.95),
    metrics = "accuracy"
  )
)

## create validation sets for model development
train_fishNum <- unique(train_seq_data$fishNum)
# lstm_val_fishNum <- sample(train_fishNum,
#                            size = floor(0.2 * length(train_fishNum)))
lstm_val_fishNum <- (
  train_std_data
  |> distinct(species, fishNum)  # Get unique species-fishNum combinations
  |> group_by(species)
  |> slice_sample(prop = 0.2)
  |> pull(fishNum)
)

## split data using fishNum
train_lstm_idx <- which(!train_seq_data$fishNum %in% lstm_val_fishNum)
val_lstm_idx <- which(train_seq_data$fishNum %in% lstm_val_fishNum)

# train_lstm_idx <- createDataPartition(
#   train_seq_data$species,
#   p = 0.9,
#   list = FALSE
# )
# val_lstm_idx <- setdiff(1:length(train_seq_data$species), train_lstm_idx)

class_counts <- table(train_seq_data$species)
total <- sum(class_counts)
class_weights <- list(
  "0" = total / (num_classes * class_counts[1]),  # LakeTrout
  "1" = total / (num_classes * class_counts[2]),  # LakeWhitefish
  "2" = total / (num_classes * class_counts[3])   # smallmouthBass
)

# callbacks_list <- list(
#   # Early stopping with shorter patience
#   callback_early_stopping(
#     monitor = "val_loss",
#     patience = 7,
#     restore_best_weights = TRUE
#   ),
#   
#   # More gradual learning rate reduction
#   callback_reduce_lr_on_plateau(
#     monitor = "val_loss",
#     factor = 0.5,  # Reduce by half instead of by 90%
#     patience = 3,
#     min_lr = 1e-6
#   )
# )
# 
# confusion_matrix_callback <- callback_lambda(
#   on_epoch_end = function(epoch, logs) {
#     # Only calculate every N epochs to save time
#     if (epoch %% 5 == 0) {  # Calculate every 5 epochs
#       # Get predictions on validation data
#       val_pred <- predict(lstm_model, train_seq_array[val_lstm_idx, , ])
#       val_pred_classes <- max.col(val_pred) - 1
#       
#       # Get true classes
#       val_true_classes <- max.col(train_lstm_y[val_lstm_idx, ]) - 1
#       
#       # Create confusion matrix
#       cm <- table(Predicted = val_pred_classes, Actual = val_true_classes)
#       
#       # Print to console
#       cat("\nEpoch", epoch, "Confusion Matrix:\n")
#       print(cm)
#       
#       # Calculate per-class metrics
#       n <- sum(cm)
#       diag <- diag(cm)
#       precision <- diag / colSums(cm)
#       recall <- diag / rowSums(cm)
#       f1 <- 2 * precision * recall / (precision + recall)
#       
#       cat("\nPrecision:", precision, "\n")
#       cat("Recall:", recall, "\n")
#       cat("F1 Score:", f1, "\n\n")
#     }
#   }
# )

# Add this callback to your callbacks list
callbacks_list <- c(callbacks_list, list(confusion_matrix_callback))


## sample weights
# Create weights based on sequence length (giving more weight to shorter sequences)
sequence_lengths <- map_dbl(train_seq_data$sequences, ~ nrow(.))

# Option 1: Inverse weighting (shorter sequences get more weight)
sample_weights <- 1 / (sequence_lengths + 1)  # Adding 1 to avoid division by zero

# Option 2: Normalize weights to sum to n_samples
sample_weights <- sample_weights * (length(sample_weights) / sum(sample_weights))

# Select only weights for training samples
train_sample_weights <- sample_weights[train_lstm_idx]

## train model
lstm_history <- (
  lstm_model
  |> fit(
    x = train_seq_array[train_lstm_idx, , ],
    y = train_lstm_y[train_lstm_idx, ],
    epochs = 100,
    batch_size = 32,
    validation_data = list(
      train_seq_array[val_lstm_idx, , ],
      train_lstm_y[val_lstm_idx, ]
    ),
    callbacks = list(
      callback_early_stopping(patience = 10),
      callback_reduce_lr_on_plateau(factor = 0.1, patience = 5)
    ),
    # callbacks = callbacks_list,
    # sample_weight = train_sample_weights,
    class_weight = class_weights,
    verbose = 1
  )
)
```


```{r}
test_seq_array
```


```{r}
## Evaluation
# Convert predictions to labels
get_pred_labels <- function(pred_probs) {
  pred_indices <- k_argmax(pred_probs) %>% as.numeric()
  all_species[pred_indices + 1]
}

## predict with test set
test_lstm_probs <- predict(lstm_model, test_seq_array)
test_lstm_preds <- test_lstm_probs |> get_pred_labels()
test_lstm_probs
test_lstm_preds

## create tibble to store predictions for model performance comparison
lstm_predictions <- tibble(
  fishNum = test_seq_data$fishNum,
  species = test_seq_data$species,
  pred_class = test_lstm_preds
)
lstm_fish_level_pred <- (
  lstm_predictions
  |> mutate(
    species = factor(species, levels = all_species),
    pred_class = factor(pred_class, levels = all_species)
  )
)

# Compute accuracy
lstm_accuracy <- (
  lstm_fish_level_pred
  |> accuracy(truth = species, estimate = pred_class)
)

## confusion matrix
lstm_conf_mat <- (
  lstm_fish_level_pred
  |> conf_mat(truth = species, estimate = pred_class)
)
  
## Output
## confusion matrix
cat("\nConfusion Matrix:\n")
print(lstm_conf_mat)

cat("\nModel Performance Metrics:\n")
print(lstm_accuracy)
```


```{r}
## create tibble to store predictions for model performance comparison
dnn_predictions <- tibble(
  fishNum = test_std_data$fishNum,
  species = test_std_data$species,
  pred_class = dnn_pred_classes
)

## count majority predictions on ping data for a single fish
```


```{r}
test_seq_data$species
```

```{r}
(
  train_std_data
  |> group_by(species)
  |> summarise(fishNum = unique(fishNum))
  # |> data.frame(fishNum = _)
  |> slice_sample(prop = 0.3)
  |> pull(fishNum)
)
```

```{r}
(
  train_std_data
  |> distinct(species, fishNum)  # Get unique species-fishNum combinations
  |> group_by(species)
  |> slice_sample(prop = 0.3)
  |> pull(fishNum)
)
```


```{r}
train_seq_array <- (
  train_seq_data$sequences
  |> map(function(x) pad_sequences(x, train_ping_max_length))
  |> unname()
  |> abind::abind(along = 3)
  |> aperm(c(3, 1, 2))
)
```

```{r}
(
  train_seq_data$sequences
  |> map(function(x) pad_sequences(x, train_ping_max_length))
  |> abind::abind(along = 3)
)
```

```{r}
one <- pad_sequences(train_seq_data$sequences, train_ping_max_length)
map(train_seq_data$sequences, one) |> dim()
```


```{r}
length(train_seq_data$sequences)
```


```{r}
sequence_info <- data.frame(
  fishNum = train_seq_data$fishNum,
  species = train_seq_data$species,
  num_rows = map_dbl(train_seq_data$sequences, nrow),
  num_cols = map_dbl(train_seq_data$sequences, ncol)
)
sequence_info
```






































































