---
title: "01_Characterstics"
knit: (function(input, encoding) {
  rmarkdown::render(input = input,
                    output_dir = here::here("Output", "HTML"),
                    knit_root_dir = rprojroot::find_rstudio_root_file())})
output:
  html_document: 
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document: 
    toc: yes
date: "2025-03-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(reshape2)
```


## Load

Loading Pre-processed data provided

```{r}
## load
load(here::here("Data", "ProcessedData", "processed_AnalysisData.Rdata"))
```

```{r}
processed_data
```

```{r}
## variable names in data
names(processed_data)
```

## Basic Exploration

### Basic Characterstics

Number of total fish species.

```{r}
unique(processed_data$species)
```

Number of fish sample in each fish species.

```{r}

## "1" been considered as Male
## "2" been considered as Female
(
  processed_data
  |> group_by(species)
  |> summarize(n = n_distinct(fishNum),
               M = n_distinct(fishNum[sex == "1"]),
               F = n_distinct(fishNum[sex == "2"]))
)
```

Create a basic characteristics table

```{r}
character_data <- (
  processed_data
  |> select(1:58)
  |> group_by(species, fishNum)
  |> slice(1)
  |> ungroup()
)
character_data
```

Separate into three dataset for each species

```{r}
LakeTrout <- processed_data[processed_data$species == "lakeTrout", ]
LakeWhiteFish <- processed_data[processed_data$species == "lakeWhitefish", ]
SmallmouthBass <- processed_data[processed_data$species == "smallmouthBass", ]
```


### Weight

Individual fish weight in each species.

```{r}
## LakeTrout, LakeWhiteFish, SmallmouthBass
weight_data = LakeTrout
(
  ggplot(weight_data, aes(x = fishNum, y = weight))
  + geom_point(aes(color = sex), size = 3, alpha = 0.7)
  + geom_line(aes(group = fishNum), alpha = 0.3)
  + scale_color_manual(values = c("1" = "blue", "2" = "red"), 
                        labels = c("1" = "Male", "2" = "Female"),
                        name = "Sex")
  + labs(
      title = paste0("Indiviaul Weight distribution in ", weight_data$species),
      x = "Fish ID",
      y = "Weight"
    )
  + theme_minimal()
  + theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "top"
    )
)
```


Weight distribution comparion between three species.

```{r}
(
  ggplot(character_data, aes(x = species, y = weight, fill = species))
  + geom_violin(alpha = 0.7, trim = FALSE)
  + geom_boxplot(width = 0.1, fill = "white", alpha = 0.7)
  + geom_jitter(aes(color = factor(sex)), width = 0.1, alpha = 0.6, size = 2)
  + scale_color_manual(values = c("1" = "blue", "2" = "red"), 
                       labels = c("1" = "Male", "2" = "Female"),
                       name = "Sex")
  + scale_fill_brewer(palette = "Set2")
  + scale_x_discrete(labels = c("lakeTrout" = "Lake Trout", 
                                "lakeWhitefish" = "Lake White Fish",
                                "smallmouthBass" = "Smallmouth Bass"))
  + labs(
      title = "Weight Distribution by Fish Species",
      x = "Species",
      y = "Weight",
      fill = "Species"
    )
  + theme_minimal()
  + theme(
      legend.position = "right",
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
)
```


### Total Length

Individual fish total length in each species.

```{r}
## LakeTrout, LakeWhiteFish, SmallmouthBass
length_data = SmallmouthBass
(
  ggplot(length_data, aes(x = fishNum, y = totalLength))
  + geom_point(aes(color = sex), size = 3, alpha = 0.7)
  + geom_line(aes(group = fishNum), alpha = 0.3)
  + scale_color_manual(values = c("1" = "blue", "2" = "red"), 
                        labels = c("1" = "Male", "2" = "Female"),
                        name = "Sex")
  + labs(
      title = paste0("Individual Total Length distribution in ", length_data$species),
      x = "Fish ID",
      y = "Total Length"
    )
  + theme_minimal()
  + theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "top"
    )
)
```


Total Length distribution comparison between three species.

```{r}

(
  ggplot(character_data, aes(x = species, y = totalLength, fill = species))
  + geom_violin(alpha = 0.7, trim = FALSE)
  + geom_boxplot(width = 0.1, fill = "white", alpha = 0.7)
  + geom_jitter(aes(color = factor(sex)), width = 0.1, alpha = 0.6, size = 2)
  + scale_color_manual(values = c("1" = "blue", "2" = "red"), 
                       labels = c("1" = "Male", "2" = "Female"),
                       name = "Sex")
  + scale_fill_brewer(palette = "Set2")
  + scale_x_discrete(labels = c("lakeTrout" = "Lake Trout", 
                                "lakeWhitefish" = "Lake White Fish",
                                "smallmouthBass" = "Smallmouth Bass"))
  + labs(
      title = "Total Length Distribution by Fish Species",
      x = "Species",
      y = "Total Length",
      fill = "Species"
    )
  + theme_minimal()
  + theme(
      legend.position = "right",
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
)
```


### Fork Length

Individual fish fork length in each species.

```{r}
## LakeTrout, LakeWhiteFish, SmallmouthBass
length_data = SmallmouthBass
(
  ggplot(length_data, aes(x = fishNum, y = forkLength))
  + geom_point(aes(color = sex), size = 3, alpha = 0.7)
  + geom_line(aes(group = fishNum), alpha = 0.3)
  + scale_color_manual(values = c("1" = "blue", "2" = "red"), 
                        labels = c("1" = "Male", "2" = "Female"),
                        name = "Sex")
  + labs(
      title = paste0("Individual Fork Length distribution in ", length_data$species),
      x = "Fish ID",
      y = "Fork Length"
    )
  + theme_minimal()
  + theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "top"
    )
)
```


Total Length distribution comparison between three species.

```{r}
(
  ggplot(character_data, aes(x = species, y = forkLength, fill = species))
  + geom_violin(alpha = 0.7, trim = FALSE)
  + geom_boxplot(width = 0.1, fill = "white", alpha = 0.7)
  + geom_jitter(aes(color = factor(sex)), width = 0.1, alpha = 0.6, size = 2)
  + scale_color_manual(values = c("1" = "blue", "2" = "red"), 
                       labels = c("1" = "Male", "2" = "Female"),
                       name = "Sex")
  + scale_fill_brewer(palette = "Set2")
  + scale_x_discrete(labels = c("lakeTrout" = "Lake Trout", 
                                "lakeWhitefish" = "Lake White Fish",
                                "smallmouthBass" = "Smallmouth Bass"))
  + labs(
      title = "Fork Length Distribution by Fish Species",
      x = "Species",
      y = "Fork Length",
      fill = "Species"
    )
  + theme_minimal()
  + theme(
      legend.position = "right",
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
)
```


### Shape difference of the caudal fin

We want to use total_length - fork_length to compute the outermost portion of the caudal fin beyond its fork. We want to see if there’s an obvious difference in the shape and the size of the caudal fin.

```{r}
character_data$caudal_fin_diff <- character_data$totalLength - character_data$forkLength
(
  ggplot(character_data, aes(x = species, y = caudal_fin_diff, fill = species))
  + geom_violin(alpha = 0.7, trim = FALSE)
  + geom_boxplot(width = 0.1, fill = "white", alpha = 0.7)
  + geom_jitter(aes(color = factor(sex)), width = 0.1, alpha = 0.6, size = 2)
  + scale_color_manual(values = c("1" = "blue", "2" = "red"), 
                       labels = c("1" = "Male", "2" = "Female"),
                       name = "Sex")
  + scale_fill_brewer(palette = "Set2")
  + scale_x_discrete(labels = c("lakeTrout" = "Lake Trout", 
                                "lakeWhitefish" = "Lake White Fish",
                                "smallmouthBass" = "Smallmouth Bass"))
  + labs(
      title = "Fork Length Distribution by Fish Species",
      x = "Species",
      y = "Fork Length",
      fill = "Species"
    )
  + theme_minimal()
  + theme(
      legend.position = "right",
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
)
```


### Weight-Length Curves

```{r}
total_length_lm <- lm(log(weight) ~ log(totalLength), data = character_data)
summary(total_length_lm)
```


```{r}
total_length_a <- exp(coef(total_length_lm)[1])
total_length_b <- coef(total_length_lm)[2]
ggplot(character_data, aes(x = totalLength, y = weight, color = species)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", formula = y ~ total_length_a * I(x^total_length_b), se = FALSE) +
  labs(title = "Weight-Length Relationship",
       x = "Length (units)",
       y = "Weight (units)") +
  facet_wrap(~ species, scales = "free")
```

```{r}
## linear regression models for weight-length parameters
wl_models <- by(character_data, character_data$species,
                function(data) {
                  model <- lm(log(weight) ~ log(totalLength), data = data)
                  a <- exp(coef(model)[1])
                  b <- coef(model)[2]
                  return(list(a = a, b = b))
                })

## parameters for each species
species_list <- levels(factor(character_data$species))
total_length_a <- sapply(species_list, 
                         function(sp){
  wl_models[[sp]]$a
})
total_length_b <- sapply(species_list, 
                         function(sp){
  wl_models[[sp]]$b
})
names(total_length_a) <- species_list
names(total_length_b) <- species_list

## predict standard w-l curves
predict_res <- data.frame()
## for loop to compute
for (sp in species_list) {
  ## find length range for each species
  length_range <- seq(
    min(character_data$totalLength[character_data$species == sp]),
    max(character_data$totalLength[character_data$species == sp]),
    length.out = 100
  )
  
  ## create standard weight for each species
  temp <- data.frame(
    totalLength = length_range,
    weight = total_length_a[sp] * length_range^total_length_b[sp],
    species = sp
  )
  
  predict_res <- rbind(predict_res, temp)
}

## plot
(
  ggplot(character_data, aes(x = totalLength, y = weight, color = species))
  + geom_point(alpha = 0.7)
  + geom_line(data = predict_res, aes(x = totalLength, y = weight, color = species), linewidth = 1)
  + labs(title = "Weight-Length Relationship",
         x = "Length (mm)",
         y = "Weight (g)")
  + facet_wrap(~ species, scales = "free")
)
```

### Dorso Lateral Height

Individual fish dorso lateral height in each species.

```{r}
## LakeTrout, LakeWhiteFish, SmallmouthBass
length_data = LakeWhiteFish
(
  ggplot(length_data, aes(x = fishNum, y = dorsoLatHeight))
  + geom_point(aes(color = sex), size = 3, alpha = 0.7)
  + geom_line(aes(group = fishNum), alpha = 0.3)
  + scale_color_manual(values = c("1" = "blue", "2" = "red"), 
                        labels = c("1" = "Male", "2" = "Female"),
                        name = "Sex")
  + labs(
      title = paste0("Indiviaul Dorso Lateral Height distribution in ", length_data$species),
      x = "Fish ID",
      y = "Dorso Lateral Height"
    )
  + theme_minimal()
  + theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "top"
    )
)
```


Dorso lateral height distribution comparison between three species.

```{r}
(
  ggplot(character_data, aes(x = species, y = dorsoLatHeight, fill = species))
  + geom_violin(alpha = 0.7, trim = FALSE)
  + geom_boxplot(width = 0.1, fill = "white", alpha = 0.7)
  + geom_jitter(aes(color = factor(sex)), width = 0.1, alpha = 0.6, size = 2)
  + scale_color_manual(values = c("1" = "blue", "2" = "red"), 
                       labels = c("1" = "Male", "2" = "Female"),
                       name = "Sex")
  + scale_fill_brewer(palette = "Set2")
  + scale_x_discrete(labels = c("lakeTrout" = "Lake Trout", 
                                "lakeWhitefish" = "Lake White Fish",
                                "smallmouthBass" = "Smallmouth Bass"))
  + labs(
      title = "Dorso Lateral Height Distribution by Fish Species",
      x = "Species",
      y = "Dorso Lateral Height",
      fill = "Species"
    )
  + theme_minimal()
  + theme(
      legend.position = "right",
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
)
```

















































