---
title: "02_Frequencies"
knit: (function(input, encoding) {
  # rmarkdown::render(input = input,
  #                   output_format = "html_document",
  #                   output_dir = here::here("Output", "HTML"),
  #                   knit_root_dir = rprojroot::find_rstudio_root_file())})
  rmarkdown::render(input = input,
                    output_format = "pdf_document",
                    output_dir = here::here("Output", "PDF"),
                    knit_root_dir = rprojroot::find_rstudio_root_file())})
output:
  html_document: 
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document: 
    toc: yes
date: "2025-03-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r warning=FALSE, message=FALSE}
library(CCA)
library(dplyr)
library(ggplot2)
library(reshape2)
library(ggridges)
library(gridExtra)
library(patchwork)
library(FactoMineR)
```


## Load

Loading Pre-processed data provided

```{r}
## load
load(here::here("Data", "ProcessedData", "processed_AnalysisData.Rdata"))
```

```{r}
processed_data
```

```{r}
## variable names in data
names(processed_data)
```


Create a dataframe only containing frequencies.
```{r}
frequency_data <- (
  processed_data
  |> select(1, 5, 59:481)
)
frequency_data |> head()
```



## Basic Frequencies

### Data cleaning

Since we notice the data of F90 to F170 are missing, we want to remove these frequencies columns from our dataset.
```{r}
## removing columns 93(F90) to 253(F170)
frequency_data <- frequency_data |> select(-c(93:253))
## name(frequency_data)
```

Separate into three dataset for each species

```{r}
LakeTrout <- frequency_data[frequency_data$species == "lakeTrout", ]
LakeWhiteFish <- frequency_data[frequency_data$species == "lakeWhitefish", ]
SmallmouthBass <- frequency_data[frequency_data$species == "smallmouthBass", ]
```


### TS Mean Between Fish Species

We want to use the mean of target strength across all ping times of each fish across all frequencies to explore the dataset by fish species. We first create a dataframe that contains all mean values as above.

```{r}
ts_mean_allTime_data <- (
  frequency_data
  |> group_by(fishNum, species)
  |> mutate(species = recode(species,
                             "lakeTrout" = "Lake Trout",
                             "lakeWhitefish" = "Lake Whitefish",
                             "smallmouthBass" = "Smallmouth Bass"))
  |> summarize(across(starts_with("F"), ~mean(., na.rm = TRUE), .names = "{.col}_mean"))
)
ts_mean_allTime_data |> head()
```

We then melt the wide format dataframe into a long format dataframe for easy visualization.

```{r}
ts_mean_allTime_data_long <- (
  ts_mean_allTime_data
  |> melt(
    id.vars = c("fishNum", "species"),
    variable.name = "Frequency",
    value.name = "TS_mean_allTime"
  )
)

ts_mean_allTime_data_long$Frequency <- as.numeric(gsub("F([0-9.]+)_mean", "\\1", 
                                                       ts_mean_allTime_data_long$Frequency))
ts_mean_allTime_data_long <- ts_mean_allTime_data_long |> arrange(fishNum, Frequency)
ts_mean_allTime_data_long |> head()
```


#### Plots

We want to show different type of plots to discover the frequency patterns of each fish species.

First we want to plot an frequency response plot showing the mean target strength across different frequencies for different fish species with each fish response in the background.

```{r}
(
  ggplot(ts_mean_allTime_data_long, aes(x = Frequency, y = TS_mean_allTime, color = species))
  ## Indiviual fish (dotted lines)
  + geom_line(
    aes(group = interaction(fishNum, species)), 
    linetype = "dotted", 
    alpha = 0.3, 
    linewidth = 0.5
  )
  ## Mean trends (bold lines)
  + stat_summary(
    fun = mean, 
    geom = "line", 
    aes(group = species), 
    linewidth = 1.2
  )
  ## Aesthetics
  + scale_x_continuous(breaks = seq(40, 260, by = 20)) 
  + labs(
    x = "Frequency (Hz)", 
    y = "Mean Target Strength (dB)", 
    color = "Fish Type"
  )
  + theme(legend.position = "top")
)
```


Plot to observe the range of frequncies for each individual fish for each species.

```{r}
(
  ggplot(ts_mean_allTime_data_long, aes(x = Frequency, y = TS_mean_allTime, color = species))
  + stat_summary(fun.data = mean_cl_normal, geom = "ribbon", alpha = 0.2, aes(fill = species))
  + stat_summary(fun = mean, geom = "line", linewidth = 1.2)
)
```


Plot to observe the range of frequencies of each species separtely.

```{r}
(
  ggplot(ts_mean_allTime_data_long, aes(x = Frequency, y = TS_mean_allTime))
  + geom_line(aes(group = fishNum), linetype = "dotted", alpha = 0.3)
  + stat_summary(fun = mean, geom = "line", color = "red", linewidth = 1.2)
  + facet_wrap(~species)
)
```


Density contour plot to show target strength measurements across differnt frequency ranges for each fish species.

```{r}
## function to create density contour plot
create_frequency_density_plot <- function(data, title){
#' Create Target Strength Frequency Response Plots
#'
#' This function creates visualization of acoustic target strength data across frequency ranges for fish.
#' The plot combines multiple visualization elements:
#'   - Hexagonal binning to show data density distribution
#'   - Individual fish measurements as dotted gray lines
#'   - Mean target strength trend line in red
#'   
#' @param data A long format data frame containing columns in this order: fishNum, species, Frequency, and TS_mean
#' @param title String for the plot title

  colnames(data)[4] <- "TS_mean"

  (
    ggplot(data, aes(x = Frequency, y = TS_mean))
    ## hexbin density layer
    + geom_hex(aes(fill = after_stat(count)), bins = 20, alpha = 0.7)
    ## Indiviiual fish lines
    + geom_line(
      aes(group = interaction(fishNum, species)), 
      linetype = "dotted", 
      alpha = 0.2, 
      color = "gray20"
    )
    ## mean trend
    + stat_summary(
      fun = mean, 
      geom = "line", 
      color = "red", 
      linewidth = 1.2
    )
    ## facet by fish type
    + facet_wrap(~species, nrow = 1)
    ## Aesthetics
    + scale_fill_viridis_c(name = "Data Density")
    + scale_x_continuous(breaks = seq(0, 260, by = 20))
    + labs(title = title, x = "Frequency (Hz)", y = "TS_mean (dB)")
    # + theme_minimal()
    + theme(legend.position = "bottom")
  )
}
```

```{r}
## split data into two frequency groups
ts_mean_allTime_low_long <- ts_mean_allTime_data_long |> filter(Frequency >= 45 & Frequency <= 89.5)
ts_mean_allTime_high_long <- ts_mean_allTime_data_long |> filter(Frequency >= 173 & Frequency <= 260)

# Create plots for both frequency ranges
plot_frequency_density_low <- create_frequency_density_plot(ts_mean_allTime_low_long, "Frequency Range: 45–89.5 Hz")
plot_frequency_density_high <- create_frequency_density_plot(ts_mean_allTime_high_long, "Frequency Range: 173–260 Hz")

# Arrange side-by-side
grid.arrange(plot_frequency_density_low, plot_frequency_density_high, ncol = 2)
```


Create plot of density distribution of target strength by fish species.

```{r}
(
  ggplot(ts_mean_allTime_data_long, aes(
    x = TS_mean_allTime,       # Numeric variable for density
    y = species,     # Categorical variable (fish species)
    fill = species   # Color by species
  ))
  ## density ridges
  + geom_density_ridges(alpha = 0.6, scale = 1.5)
  + labs(
    title = "Density Distribution of TS_mean by Fish Species",
    x = "Mean Target Strength (dB)",
    y = "Fish Species"
  )
  + theme_minimal()
  + scale_fill_manual(values = c("#1f77b4", "#ff7f0e", "#2ca02c"))
)
```

We create another density contour plot and we want to observe if there any potential dominant frequency response region by each fish species.

```{r warning=FALSE}
(
  ggplot(ts_mean_allTime_data_long, aes(x = Frequency, y = TS_mean_allTime, color = species))
  ## density contours
  + geom_density_2d(aes(fill = species), alpha = 0.2, contour_var = "density")
  ## frequency response trend lines
  + geom_smooth(
    method = "loess",
    formula = y ~ x,
    se = FALSE,  # Remove confidence bands
    linewidth = 1.2
  )
  ## Aesthetics
  + labs(
    title = "Dominant Regions and Frequency Response by Fish Species",
    x = "Frequency (kHz)",
    y = "Mean Target Strength (dB)",
    color = "Fish Species"
  )
  + scale_color_manual(values = c("#1f77b4", "#ff7f0e", "#2ca02c"))
  + scale_fill_manual(values = c("#1f77b4", "#ff7f0e", "#2ca02c"))
  + theme(legend.position = "bottom")
  + theme_minimal()
)
```


### TS Mean In Fish Species

This section we want to investigate the potential trend / pattern in the frequency response in an indiviual fish species.

#### Lake Trout

We first want to investigate the frequency response across different ping time for a single fish. We want to focus on `fishNum = LT001`.

```{r}
LT001_frequency <- LakeTrout[LakeTrout$fishNum == "LT004", ]
# LT001_frequency <- LakeWhiteFish[LakeWhiteFish$fishNum == "LWF003", ]
# LT001_frequency <- SmallmouthBass[SmallmouthBass$fishNum == "SMB005", ]
```

```{r}
## melt dataframe into plottable format
LT001_frequency$ping_id <- 1:nrow(LT001_frequency)
LT001_frequency_long <- melt(
  LT001_frequency,
  id.vars = c("fishNum", "species", "ping_id"),
  variable.name = "Frequency",
  value.name = "TS"
)
LT001_frequency_long$Frequency <- as.numeric(gsub("F", "", 
                                                  LT001_frequency_long$Frequency))
LT001_frequency_long |> head()
```

Plot to create visualization of frequency response of fish `LT001` across ping time. 

```{r}
(
  ggplot(LT001_frequency_long, aes(x = Frequency, y = TS))
  ## Indiviual ping time response as dotted lines
  + geom_line(aes(group = ping_id), linetype = "dotted", alpha = 0.3, color = "gray40")
  ## mean respinse as solid line
  + stat_summary(fun = mean, geom = "line", color = "red", linewidth = 1.2)
  + stat_summary(fun = median, geom = "line", color = "blue", linewidth = 1.2)
  ## confidence interval for mean
  # + stat_summary(fun.data = mean_cl_normal, geom = "ribbon", alpha = 0.2, fill = "red")
  + labs(
    title = "Acoustic Frequency Response",
    subtitle = "Individual ping responses with mean trend",
    x = "Frequency (Hz)",
    y = "Target Strength (dB)"
  )
  + theme(
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )
  + theme_minimal()
)
```

Next, we want to compte the difference between the frequency response from the mean from different ping time.

```{r}
## compute mean
LT001_frequency_mean <- (
  LT001_frequency_long
  |> group_by(Frequency)
  |> summarize(TS_mean = mean(TS, na.rm = TRUE))
)

## left join and compute difference
LT001_frequency_long <- (
  LT001_frequency_long
  |> left_join(LT001_frequency_mean, by = "Frequency")
  |> mutate(diff_from_mean = TS - TS_mean)
)
LT001_frequency_long |> head()
```

```{r}
(
  ggplot(LT001_frequency_long, aes(x = Frequency, y = diff_from_mean))
  ## Indiviual ping time response as dotted lines
  + geom_line(aes(group = ping_id), linetype = "dotted", alpha = 0.3, color = "gray40")
  ## mean respinse as solid line
  + stat_summary(fun = mean, geom = "line", color = "red", linewidth = 1.2)
  + stat_summary(fun = median, geom = "line", color = "blue", linewidth = 1.2)
  ## confidence interval for mean
  # + stat_summary(fun.data = mean_cl_normal, geom = "ribbon", alpha = 0.2, fill = "red")
  + labs(
    title = "Acoustic Frequency Response",
    subtitle = "Individual ping responses with mean trend",
    x = "Frequency (Hz)",
    y = "Target Strength (dB)"
  )
  + theme(
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )
  + theme_minimal()
)
```

```{r}
# Calculate confidence intervals for the differences from mean
ci_data <- LT001_frequency_long %>%
  group_by(Frequency) %>%
  summarize(
    mean_diff = mean(diff_from_mean, na.rm = TRUE),
    sd_diff = sd(diff_from_mean, na.rm = TRUE),
    n = n(),
    # Calculate 95% confidence interval
    ci_lower = mean_diff - qt(0.975, n-1) * sd_diff / sqrt(n),
    ci_upper = mean_diff + qt(0.975, n-1) * sd_diff / sqrt(n),
    # Calculate variability metrics
    cv = sd_diff / abs(mean_diff + 0.0001), # Coefficient of variation (adding small constant to avoid div by 0)
    range = max(diff_from_mean, na.rm = TRUE) - min(diff_from_mean, na.rm = TRUE)
  )

# Create a visualization of the confidence intervals
ggplot(ci_data, aes(x = Frequency)) +
  # Add confidence interval as ribbon
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.3, fill = "blue") +
  # Add mean line
  geom_line(aes(y = mean_diff), color = "blue", linewidth = 1) +
  # Add zero reference line
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  # Add variability plot
  # geom_line(aes(y = cv), color = "green", linewidth = 1) +
  # Create second y-axis for coefficient of variation
  scale_y_continuous(
    name = "Difference from Mean (dB)",
    sec.axis = sec_axis(~ ., name = "Coefficient of Variation")
  ) +
  labs(
    title = paste("Variation in Acoustic Response for", unique(LT001_frequency_long$fishNum)),
    subtitle = "Blue: 95% CI of differences from mean, Green: Coefficient of Variation",
    x = "Frequency (Hz)"
  ) +
  theme_minimal() +
  theme(
    axis.title.y.right = element_text(color = "green"),
    axis.text.y.right = element_text(color = "green")
  )
```


#### Principal Component Analysis

We want to perform PCA in Lake Trout and we want to see if specified frequencies is contributing strongly to the variance of Lake Trout.

```{r}
## aggregating dataframe into mean frquencies by each fish and scale
LakeTrout_agg <- (
  LakeTrout
  |> group_by(fishNum)
  |> filter(fishNum != "LT008") ## contains missing bal
  |> summarise(across(starts_with("F"), mean, na.rm = TRUE))
  |> ungroup()
  |> dplyr::select(-fishNum)
  |> scale()
)

LakeTrout_pca <- PCA(LakeTrout_agg, graph = FALSE)
LakeTrout_pca$eig 
```

```{r}
LakeTrout_loadings_pc1 <-(
  LakeTrout_pca$var$coord[, 1]
  |> as.data.frame()
  |> rename(Loading = "LakeTrout_pca$var$coord[, 1]")
  |> mutate(Frequency = rownames(LakeTrout_pca$var$coord))
  |> arrange(desc(abs(Loading)))
)
# loadings_pc1
LakeTrout_loadings_pc1_top_frequencies <- LakeTrout_loadings_pc1 |> head(20)
print(LakeTrout_loadings_pc1_top_frequencies)
```

```{r}
ggplot(LakeTrout_loadings_pc1_top_frequencies, aes(x = reorder(Frequency, -abs(Loading)), y = Loading)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Top Frequencies Driving Variability in LakeTrout",
       x = "Frequency (kHz)", y = "PC1 Loading") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
cor_mat <- cor(LakeTrout_agg) |> as.data.frame()
cor_mat |> head()
```



### Lake White Fish

#### PCA

```{r}
is.na(LakeWhiteFish) |> sum()
```


```{r}
## aggregating dataframe into mean frquencies by each fish and scale
LakeWhiteFish_agg <- (
  LakeWhiteFish
  |> group_by(fishNum)
  |> summarise(across(starts_with("F"), mean, na.rm = TRUE))
  |> ungroup()
  |> dplyr::select(-fishNum)
  |> scale()
)

LakeWhiteFish_pca <- PCA(LakeWhiteFish_agg, graph = FALSE)
LakeWhiteFish_pca$eig
```

```{r}
LakeWhiteFish_loadings_pc1 <-(
  LakeWhiteFish_pca$var$coord[, 1]
  |> as.data.frame()
  |> rename(Loading = "LakeWhiteFish_pca$var$coord[, 1]")
  |> mutate(Frequency = rownames(LakeWhiteFish_pca$var$coord))
  |> arrange(desc(abs(Loading)))
)
# loadings_pc1
LakeWhiteFish_loadings_pc1_top_frequencies <- LakeWhiteFish_loadings_pc1 |> head(20)
print(LakeWhiteFish_loadings_pc1_top_frequencies)
```


#### Smallmouth Bass

```{r}
is.na(SmallmouthBass) |> sum()
```


```{r}
## aggregating dataframe into mean frquencies by each fish and scale
SmallmouthBass_agg <- (
  SmallmouthBass
  |> group_by(fishNum)
  |> summarise(across(starts_with("F"), mean, na.rm = TRUE))
  |> ungroup()
  |> dplyr::select(-fishNum)
  |> scale()
)

SmallmouthBass_pca <- PCA(SmallmouthBass_agg, graph = FALSE)
SmallmouthBass_pca$eig
```

```{r}
SmallmouthBass_loadings_pc1 <-(
  SmallmouthBass_pca$var$coord[, 1]
  |> as.data.frame()
  |> rename(Loading = "SmallmouthBass_pca$var$coord[, 1]")
  |> mutate(Frequency = rownames(SmallmouthBass_pca$var$coord))
  |> arrange(desc(abs(Loading)))
)
# loadings_pc1
SmallmouthBass_loadings_pc1_top_frequencies <- SmallmouthBass_loadings_pc1 |> head(20)
print(SmallmouthBass_loadings_pc1_top_frequencies)
```

```{r}
(SmallmouthBass
  |> group_by(fishNum)
  |> summarise(across(starts_with("F"), mean, na.rm = TRUE))
  |> ungroup()
  |> dplyr::select(-fishNum)
 |> summarise(across(starts_with("F"), mean, na.rm = TRUE))
 )
```

### Overall PCA
```{r}
LT_mean_frequency <- (
  LakeTrout
  |> group_by(fishNum)
  |> filter(fishNum != "LT008")
  |> dplyr::select(matches("^F(1[7-9][0-9](\\.[0-9])?|2[0-5][0-9](\\.[0-9])?|260(\\.[0-9])?)$"))
  |> summarise(across(starts_with("F"), mean, na.rm = TRUE))
  |> ungroup()
  |> dplyr::select(-fishNum)
  |> summarise(across(starts_with("F"), mean, na.rm = TRUE))
)
LT_mean_frequency |> head()
```
```{r}
LWF_mean_frequency <- (
  LakeWhiteFish
  |> group_by(fishNum)
  |> dplyr::select(matches("^F(1[7-9][0-9](\\.[0-9])?|2[0-5][0-9](\\.[0-9])?|260(\\.[0-9])?)$"))
  |> summarise(across(starts_with("F"), mean, na.rm = TRUE))
  |> ungroup()
  |> dplyr::select(-fishNum)
  |> summarise(across(starts_with("F"), mean, na.rm = TRUE))
)
LWF_mean_frequency |> head()
```
```{r}
SB_mean_frequency <- (
  SmallmouthBass
  |> group_by(fishNum)
  |> dplyr::select(matches("^F(1[7-9][0-9](\\.[0-9])?|2[0-5][0-9](\\.[0-9])?|260(\\.[0-9])?)$"))
  |> summarise(across(starts_with("F"), mean, na.rm = TRUE))
  |> ungroup()
  |> dplyr::select(-fishNum)
  |> summarise(across(starts_with("F"), mean, na.rm = TRUE))
)
SB_mean_frequency |> head()
```


```{r}
## prepare fish species PCA comparison data
prepare_species_PCA_data <- function(species, mean_freq_df, loadings_df) {
  
  ## convery mean frequecy data to long format
  species_long <- melt(
    mean_freq_df,
    variable.name = "Frequency",
    value.name = "TS"
    
  )
  ## removing "F" from frequency and make it numerical
  species_long$Frequency <- as.numeric(gsub("F", "",
                                            species_long$Frequency))
  loadings_df$Frequency <- as.numeric(gsub("F", "",
                                            loadings_df$Frequency))
  
  ## Add species name
  species_long$Species <- species
  
  ## join with loadings
  species_combined <- merge(
    species_long,
    loadings_df[, c("Frequency", "Loading")],
    by = "Frequency",
    all.x = TRUE
  )
  
  return(species_combined)
}
```

```{r}
## prepare PCA data for each species
LakeTrout_PCA_data <- prepare_species_PCA_data(
  "Lake Trout",
  LT_mean_frequency,
  LakeTrout_loadings_pc1
)

LakeWhiteFish_PCA_data <- prepare_species_PCA_data(
  "Lake Whitefish",
  LWF_mean_frequency,
  LakeWhiteFish_loadings_pc1
)

SmallmouthBass_PCA_data <- prepare_species_PCA_data(
  "Smallmouth Bass",
  SB_mean_frequency,
  SmallmouthBass_loadings_pc1
)

all_species_PCA_data <- rbind(
  LakeTrout_PCA_data,
  LakeWhiteFish_PCA_data,
  SmallmouthBass_PCA_data
)
```

```{r}
all_species_PCA_data |> head()
```

```{r}
## create plots to show PCA laodings and TS_mean to visulize the PCA data
species_colors <- c(
  "Lake Trout" = "#1f77b4", 
  "Lake Whitefish" = "#2ca02c", 
  "Smallmouth Bass" = "#ff7f0e"
)

## PCA loading plots
all_species_PCA_plot <- (
  ggplot(all_species_PCA_data,
         aes(x = Frequency, y = Loading, color = Species))
  + geom_line(linewidth = 0.5)
  + scale_color_manual(values = species_colors)
  + labs(
    title = "PC1 Loadings by species",
    y = "PC1 Loading"
  )
  + theme(
    axis.title.x = element_blank(), 
    axis.text.x = element_blank(),  
    legend.position = "none"         
  )
  + theme_minimal()
)

## TS_mean plots
all_species_TS_mean_plot <- (
  ggplot(all_species_PCA_data,
         aes(x = Frequency, y = TS, color = Species))
  + geom_line(linewidth = 0.8)
  + scale_color_manual(values = species_colors)
  + labs(
    title = "Mean Target Strength by Species",
    y = "Target Strength",
    x = "Frequency"
  )
  + theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  
    legend.position = "bottom"                        
  )
  + theme_minimal()
)

## combined plots
all_species_PCA_combined_plot <- (
  all_species_PCA_plot / all_species_TS_mean_plot
  + plot_layout(heights = c(1, 1.2))
)
all_species_PCA_combined_plot
```














































